---
title: "flux-model-comparison"
output: html_document
date: "2025-08-08"
---

# OBJECTIVES
-In this tutorial, we will compare the tower FLUXNET tower data with the ELM output. There are 10 overlapping variables that we can compare between the two data sets. Check the "Crosswalk table" for these variables. 
  
-There are 4 parts to this tutorial. Parts 1 to 3 will create time-series graphs for comparing environmental variables, energy exchange variables, and carbon flux variables. In Part 4, we use a simple tool to discern the level of agreement between flux tower data and ELM output. 
  

  
# IMPORTANT NOTES
- This script needs access to Gmail to download the data from Google drive

- In each part, there is a "USER INPUT" section where you should change the variables to compare, units, conversion factors (refer to the crosswalk variable table), years of data, and graph title accordingly

-This tutorial is also used for comparing the adjusted and default simulations to flux tower data 
  
  
  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lubridate)
library(data.table)
library(ggplot2)
library(ggpubr)
library(tidyverse)
library (rlang)
library(ncdf4) # to read nc data
library(RColorBrewer)
library (purrr)

my_theme <- theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 12),
    # panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = "white"),
    plot.background = element_rect(fill = "white", color = "white")
  )
```

#USER INPUT SECTION
-change site name and URLs according to your study site
```{r}
# change to the ID of your study site here
site_name <- "US_ARM" 

# flux tower directory here (FLUXNET_SUBSET_DD file)
flux_file <- "C:/Users/ruzol/Downloads/fluxnet/AMF_US-ARM_FLUXNET_SUBSET_DD_2003-2023_4-6.csv"

# default ELM output directory here
elm_file <- "C:/Users/ruzol/Downloads/fluxnet/ELM_output_US_ARM.nc"

# second ELM output directory here
elm_v2_file <- "C:/Users/ruzol/Downloads/fluxnet/US_ARM_model_v2.nc"
  
#starting year of flux tower data
start_year <- 2000
  
# Selected variables from ELM output to be compared with flux measurement
vars_to_extract <- c(
  "NEE",          # Net ecosystem exchange of carbon
  "GPP",          # Gross primary production
  "ER",           # Ecosystem respiration
  "TSOI_10CM",    # Soil temperature at 10 cm
  "FSH",          # Sensible heat flux
  "EFLX_LH_TOT",  # Total latent heat flux
  "TSA",          # surface air temperature
  "PBOT",         # atmospheric pressure
  "RAIN",         # Precipitation
  "FGR"           # Soil heat flux
)



```


# Read data from Google drive 
```{r}
# READ FLUX DATA
{
  df.flux <- fread(flux_file)
  # covert -9999 to NA
  df.flux <- df.flux %>%
    mutate(across(everything(), ~na_if(. , -9999)))
  df.flux <- df.flux %>% # add more time variables
    mutate(
      TIMESTAMP = ymd(TIMESTAMP),
      year = year(TIMESTAMP),
      month = month(TIMESTAMP),
      date = as.Date(TIMESTAMP),
      doy = yday(TIMESTAMP)
    )
}

# READ DEFAULT ELM OUTPUT
{
  nc <- nc_open(elm_file) # Open the NetCDF file
  
  # Extract time (mcdate)
  mcdate <- ncvar_get(nc, "mcdate")
  data_list <- lapply(vars_to_extract, function(var) ncvar_get(nc, var))
  
  # Combine into data frame
  df.model <- data.frame(mcdate = mcdate)
  for (i in seq_along(vars_to_extract)) {
    df.model[[vars_to_extract[i]]] <- data_list[[i]]
  }
  
  # Close NetCDF file
  nc_close(nc)
  
  # modify time variables
  df.model <- df.model %>%
    mutate(
      mcdate = ymd(mcdate),
      year = year(mcdate),
      month = month(mcdate),
      date = as.Date(mcdate),
      doy = yday(mcdate)
    ) %>%
    rename(TIMESTAMP = mcdate)
}

# READ ADJUSTED ELM OUTPUT
{
  nc_adj <- nc_open(elm_v2_file) # Open the NetCDF file
  
  
  # Extract time (mcdate)
  mcdate <- ncvar_get(nc_adj, "mcdate")
  data_list <- lapply(vars_to_extract, function(var) ncvar_get(nc_adj, var))
  
  # Combine into data frame
  df.model_adj <- data.frame(mcdate = mcdate)
  for (i in seq_along(vars_to_extract)) {
    df.model_adj[[vars_to_extract[i]]] <- data_list[[i]]
  }
  
  # Close NetCDF file
  nc_close(nc_adj)
  
  # modify time variables
  df.model_adj <- df.model_adj %>%
    mutate(
      mcdate = ymd(mcdate),
      year = year(mcdate),
      month = month(mcdate),
      date = as.Date(mcdate),
      doy = yday(mcdate)
    ) %>%
    rename(TIMESTAMP = mcdate)
}

# combine flux data and model output
df.model = df.model[df.model$year >= start_year, ] # please change to the start year of flux measurement for you study site
df.model_adj = df.model_adj[df.model_adj$year >= start_year, ] # please change to the start year of flux measurement for you study site
df.model_adj <- df.model_adj %>%
  rename_with(~ paste0(., "_v2"), .cols = -TIMESTAMP)
df.combined <- df.model %>%
  left_join(df.flux, by = "TIMESTAMP") %>%
  left_join(df.model_adj, by = "TIMESTAMP")
names(df.combined)


```


#PART 1: Comparison of environmental variables  between flux tower and ELM data

- The example here is currently set for comparing air temperature
- Use the crosswalk table for the variable names, units, and conversion factors 

```{r}
# ===== USER INPUT SECTION ====================================================
flux_var <- "TA_F"     # Flux variable name
model_var <- "TSA"     # Model variable name
model_v2_var <- "TSA_v2" #Variable name from the adjusted model
convert_model_units <- function(x) x -273.15  # Set equation if needed

selected_years <- list(2000:2015) #change years to plot; it can handle several batches (ex: list(2001:2005, 2010:20015))

compare_name <- "Air Temperature"   # variable being compared
flux_label <- expression(AirTemp[flux]~(degC)) #change label name and units as needed
model_label <- expression(AirTemp[model]~(degC)) #change label name and units as needed
model_v2_label <- expression(AirTemp[model_v2]~(degC)) #change label name and units as needed
plot_title <- "Air Temp Comparison" #change plot title as needed
time_series_label <- expression(AirTemp~(degC)) # change label name and units as needed


# =============================================================================

year_range <- unlist(selected_years)

# ==== Derived comparison columns ====
flux_col <- paste0(compare_name, "_flux")
model_col <- paste0(compare_name, "_model")
model_v2_col <- paste0(compare_name, "_model_v2")

qc_col <- paste0(flux_var, "_QC")  # dynamically get QC column name

# Check if QC column is all NA
if (all(is.na(df.combined[[qc_col]]))) {
  # No QC filtering
  df.combined_new <- df.combined %>%
    mutate(
      !!flux_col := .data[[flux_var]],
      !!model_col := convert_model_units(.data[[model_var]]),
      !!model_v2_col := convert_model_units(.data[[model_v2_var]])
    ) %>%
    filter(year.x %in% year_range)
} else {
  # Apply QC filtering
  df.combined_new <- df.combined %>%
    filter(!is.na(.data[[qc_col]])) %>%   # remove NA QC
    mutate(
      !!flux_col := .data[[flux_var]],
      !!model_col := convert_model_units(.data[[model_var]]),
      !!model_v2_col := convert_model_units(.data[[model_v2_var]])
    ) %>%
    filter(year.x %in% year_range)
}

# ==== Dynamic axis limits ====
all_x_vals <- df.combined_new[[flux_col]]
all_y_vals <- c(df.combined_new[[model_col]], df.combined_new[[model_v2_col]])
x_limits <- c(floor(min(all_x_vals, na.rm = TRUE)), ceiling(max(all_x_vals, na.rm = TRUE)))
y_limits <- c(floor(min(all_y_vals, na.rm = TRUE)), ceiling(max(all_y_vals, na.rm = TRUE)))

# ==== SCATTER PLOTS ====
# Model v1 vs Flux
p1 <- ggscatter(
  data = df.combined_new,
  x = flux_col,
  y = model_col,
  color = "steelblue",
  add = "reg.line",
  add.params = list(color = "black", size = 1),
  conf.int = TRUE,
  cor.coef = TRUE,
  cor.coeff.args = list(size = 6),
  label.x = x_limits[1],
  label.y = y_limits[2],
  title = paste(plot_title, "- Model v1"),
  ggtheme = theme_pubr()
) +
  stat_regline_equation(
    aes(label = ..eq.label..),
    label.x = x_limits[1],
    label.y = y_limits[1],
    size = 6
  ) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  xlim(x_limits) + ylim(y_limits) +
  xlab(flux_label) + ylab(model_label) +
  my_theme

# Model v2 vs Flux
p2 <- ggscatter(
  data = df.combined_new,
  x = flux_col,
  y = model_v2_col,
  color = "darkgreen",
  add = "reg.line",
  add.params = list(color = "black", size = 1),
  conf.int = TRUE,
  cor.coef = TRUE,
  cor.coeff.args = list(size = 6),
  label.x = x_limits[1],
  label.y = y_limits[2],
  title = paste(plot_title, "- Model v2"),
  ggtheme = theme_pubr()
) +
  stat_regline_equation(
    aes(label = ..eq.label..),
    label.x = x_limits[1],
    label.y = y_limits[1],
    size = 6
  ) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  xlim(x_limits) + ylim(y_limits) +
  xlab(flux_label) + ylab(model_v2_label) +
  my_theme

print(p1)
print(p2)

# ==== TIME SERIES PLOT ====
df_long <- df.combined_new %>%
  pivot_longer(
    cols = all_of(c(flux_col, model_col, model_v2_col)),
    names_to = "source",
    values_to = compare_name
  ) %>%
  filter(year.x %in% year_range)

y_limits <- c(floor(min(df_long[[compare_name]], na.rm = TRUE)),
              ceiling(max(df_long[[compare_name]], na.rm = TRUE)))

all_years <- sort(unique(df_long$year.x))
for (i in seq(1, length(all_years), by = 4)) {
  years_to_plot <- all_years[i:min(i+3, length(all_years))]
  df_subset <- df_long %>% filter(year.x %in% years_to_plot)

  p <- ggplot(df_subset, aes(x = doy.x, y = .data[[compare_name]], color = source)) +
    geom_point(alpha = 0.3, size = 0.7) +
    geom_smooth(se = FALSE, method = "loess", span = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    facet_wrap(~ year.x, ncol = 2) +
    xlab("DOY") +
    ylab(time_series_label) +
    scale_color_manual(
      values = setNames(c("red", "blue", "darkgreen"), 
                        c(flux_col, model_col, model_v2_col)),
      labels = c("Flux Tower", "Model v1", "Model v2")
    ) +
    ylim(y_limits) +
    labs(title = paste0(compare_name, " Comparison")) +
    my_theme

  print(p)
}
```


# PART 2: Energy exchange variables comparison between flux tower and ELM data

```{r}
# ===== USER INPUT SECTION ====================================================
flux_var <- "LE_F_MDS"     # Flux variable name
model_var <- "EFLX_LH_TOT"     # Model variable name
model_v2_var <- "EFLX_LH_TOT_v2" #Model v2 variable name
convert_model_units <- function(x) x  # Set equation if needed

selected_years <- list(2000:2015) #change years to plot; it can handle several batches (ex: list(2001:2005, 2010:20015))

compare_name <- "Latent heat flux"   # variable being compared
flux_label <- expression(LE[flux]~(W/m^2)) #change label name and units as needed
model_label <- expression(LE[model]~(W/m^2)) #change label name and units as needed
model_v2_label <- expression(LE[model_v2]~(W/m^2)) #change label name and units as needed
plot_title <- "Latent Heat Flux Comparison" #change plot title as needed
time_series_label <- expression(LE_flux~(W/m^2)) # change label name and units as needed


# =============================================================================

year_range <- unlist(selected_years)

# ==== Derived comparison columns ====
flux_col <- paste0(compare_name, "_flux")
model_col <- paste0(compare_name, "_model")
model_v2_col <- paste0(compare_name, "_model_v2")

qc_col <- paste0(flux_var, "_QC")  # dynamically get QC column name

# Check if QC column is all NA
if (all(is.na(df.combined[[qc_col]]))) {
  # No QC filtering
  df.combined_new <- df.combined %>%
    mutate(
      !!flux_col := .data[[flux_var]],
      !!model_col := convert_model_units(.data[[model_var]]),
      !!model_v2_col := convert_model_units(.data[[model_v2_var]])
    ) %>%
    filter(year.x %in% year_range)
} else {
  # Apply QC filtering
  df.combined_new <- df.combined %>%
    filter(!is.na(.data[[qc_col]])) %>%   # remove NA QC
    mutate(
      !!flux_col := .data[[flux_var]],
      !!model_col := convert_model_units(.data[[model_var]]),
      !!model_v2_col := convert_model_units(.data[[model_v2_var]])
    ) %>%
    filter(year.x %in% year_range)
}


# ==== Dynamic axis limits for scatter plots ====
all_x_vals <- df.combined_new[[flux_col]]
all_y_vals <- c(df.combined_new[[model_col]], df.combined_new[[model_v2_col]])
x_limits <- c(floor(min(all_x_vals, na.rm = TRUE)), ceiling(max(all_x_vals, na.rm = TRUE)))
y_limits <- c(floor(min(all_y_vals, na.rm = TRUE)), ceiling(max(all_y_vals, na.rm = TRUE)))

# ==== SCATTER PLOTS ====
# Model v1 vs Flux
p1 <- ggscatter(
  data = df.combined_new,
  x = flux_col,
  y = model_col,
  color = "steelblue",
  add = "reg.line",
  add.params = list(color = "black", size = 1),
  conf.int = TRUE,
  cor.coef = TRUE,
  cor.coeff.args = list(size = 6),
  label.x = x_limits[1],
  label.y = y_limits[2],
  title = paste(plot_title, "- Model v1"),
  ggtheme = theme_pubr()
) +
  stat_regline_equation(
    aes(label = ..eq.label..),
    label.x = x_limits[1],
    label.y = y_limits[1],
    size = 6
  ) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  xlim(x_limits) + ylim(y_limits) +
  xlab(flux_label) + ylab(model_label) +
  my_theme

# Model v2 vs Flux
p2 <- ggscatter(
  data = df.combined_new,
  x = flux_col,
  y = model_v2_col,
  color = "orange",
  add = "reg.line",
  add.params = list(color = "black", size = 1),
  conf.int = TRUE,
  cor.coef = TRUE,
  cor.coeff.args = list(size = 6),
  label.x = x_limits[1],
  label.y = y_limits[2],
  title = paste(plot_title, "- Model v2"),
  ggtheme = theme_pubr()
) +
  stat_regline_equation(
    aes(label = ..eq.label..),
    label.x = x_limits[1],
    label.y = y_limits[1],
    size = 6
  ) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  xlim(x_limits) + ylim(y_limits) +
  xlab(flux_label) + ylab(model_v2_label) +
  my_theme

print(p1)
print(p2)

# ==== TIME SERIES PLOT ====
df_long <- df.combined_new %>%
  pivot_longer(
    cols = all_of(c(flux_col, model_col, model_v2_col)),
    names_to = "source",
    values_to = compare_name
  ) %>%
  filter(year.x %in% year_range)

y_limits <- c(floor(min(df_long[[compare_name]], na.rm = TRUE)),
              ceiling(max(df_long[[compare_name]], na.rm = TRUE)))

all_years <- sort(unique(df_long$year.x))
for (i in seq(1, length(all_years), by = 4)) {
  years_to_plot <- all_years[i:min(i+3, length(all_years))]
  df_subset <- df_long %>% filter(year.x %in% years_to_plot)

  p <- ggplot(df_subset, aes(x = doy.x, y = .data[[compare_name]], color = source)) +
    geom_point(alpha = 0.3, size = 0.7) +
    geom_smooth(se = FALSE, method = "loess", span = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    facet_wrap(~ year.x, ncol = 2) +
    xlab("DOY") +
    ylab(time_series_label) +
    scale_color_manual(
      values = setNames(c("red", "blue", "orange"), 
                        c(flux_col, model_col, model_v2_col)),
      labels = c("Flux Tower", "Model v1", "Model v2")
    ) +
    ylim(y_limits) +
    labs(title = paste0(compare_name, " Comparison")) +
    my_theme

  print(p)
}

```



# PART 3: Carbon Flux variables comparison between flux tower and ELM data

```{r}
# ===== USER INPUT SECTION ====================================================
flux_var <- "NEE_VUT_REF"     # Flux variable name
model_var <- "NEE"            # Model variable name
model_v2_var <- "NEE_v2"      # Model v2 variable name

convert_model_units <- function(x) x * 86400 # Set equation if needed

selected_years <- list(2000:2015) # Change years to plot; can handle several batches

compare_name <- "Net Ecosystem Exchange"   
flux_label <- expression(NEE[flux]~(gC/m^2/day)) 
model_label <- expression(NEE[model]~(gC/m^2/day)) 
model_v2_label <- expression(NEE[model~v2]~(gC/m^2/day)) 
plot_title <- "NEE Comparison" 
time_series_label <- expression(NEE~(gC/m^2/day)) 

# =============================================================================

year_range <- unlist(selected_years)

# ==== Derived comparison columns ====
flux_col <- paste0(compare_name, "_flux")
model_col <- paste0(compare_name, "_model")
model_v2_col <- paste0(compare_name, "_model_v2")
qc_col <- paste0(flux_var, "_QC")

# ==== Conditional QC filtering ====
if (qc_col %in% names(df.combined) && !all(is.na(df.combined[[qc_col]]))) {
  df.combined <- df.combined %>% filter(!is.na(.data[[qc_col]]))
}

# ==== Create comparison dataframe ====
df.combined_new <- df.combined %>%
  mutate(
    !!flux_col    := .data[[flux_var]],
    !!model_col   := convert_model_units(.data[[model_var]]),
    !!model_v2_col := convert_model_units(.data[[model_v2_var]])
  ) %>%
  filter(year.x %in% year_range)

# ==== Dynamic axis limits for scatter plot ====
x_vals <- df.combined_new[[flux_col]]
y_vals_all <- c(df.combined_new[[model_col]], df.combined_new[[model_v2_col]])
x_limits <- c(floor(min(x_vals, na.rm = TRUE)), ceiling(max(x_vals, na.rm = TRUE)))
y_limits <- c(floor(min(y_vals_all, na.rm = TRUE)), ceiling(max(y_vals_all, na.rm = TRUE)))

# ==== X-Y SCATTER PLOT ====
#flux tower vs model output
p1 <- ggscatter(
  data = df.combined_new,
  x = flux_col,
  y = model_col,
  color = "steelblue",
  add = "reg.line",
  add.params = list(color = "black", size = 1),
  conf.int = TRUE,
  cor.coef = TRUE,
  cor.coeff.args = list(size = 6),
  label.x = x_limits[1],
  label.y = y_limits[2],
  title = plot_title,
  ggtheme = theme_pubr()
) +
  stat_regline_equation(
    aes(label = ..eq.label..),
    label.x = x_limits[1],
    label.y = y_limits[1],
    size = 6
  ) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  xlim(x_limits) + ylim(y_limits) +
  xlab(flux_label) + ylab(model_label) +
  my_theme

#flux tower vs model v2 output
p2 <- ggscatter(
  data = df.combined_new,
  x = flux_col,
  y = model_v2_col,
  color = "darkgreen",
  add = "reg.line",
  add.params = list(color = "black", size = 1),
  conf.int = TRUE,
  cor.coef = TRUE,
  cor.coeff.args = list(size = 6),
  label.x = x_limits[1],
  label.y = y_limits[2],
  title = plot_title,
  ggtheme = theme_pubr()
) +
  stat_regline_equation(
    aes(label = ..eq.label..),
    label.x = x_limits[1],
    label.y = y_limits[1],
    size = 6
  ) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  xlim(x_limits) + ylim(y_limits) +
  xlab(flux_label) + ylab(model_v2_label) +
  my_theme

print (p1)
print (p2)

# ==== TIME SERIES PLOT ====
df_long <- df.combined_new %>%
  pivot_longer(
    cols = all_of(c(flux_col, model_col, model_v2_col)),
    names_to = "source",
    values_to = compare_name
  )

y_vals <- df.combined_new[[flux_col]]
y_limits <- c(floor(min(y_vals, na.rm = TRUE)), ceiling(max(y_vals, na.rm = TRUE)))

all_years <- sort(unique(df_long$year.x))
for (i in seq(1, length(all_years), by = 4)) {
  years_to_plot <- all_years[i:min(i+3, length(all_years))]
  df_subset <- df_long %>% filter(year.x %in% years_to_plot)

  p <- ggplot(df_subset, aes(x = doy.x, y = .data[[compare_name]], color = source)) +
    geom_point(alpha = 0.3, size = 0.7) +
    geom_smooth(se = FALSE, method = "loess", span = 0.2) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    facet_wrap(~ year.x, ncol = 2) +
    xlab("DOY") +
    ylab(time_series_label) +
    scale_color_manual(
      values = setNames(c("red", "blue", "darkgreen"), c(flux_col, model_col, model_v2_col)),
      labels = c("Flux", "Model", "Model v2")
    ) +
    ylim(y_limits) +
    labs(title = paste0(compare_name, " Comparison")) +
    my_theme

  print(p)
}

# ==== Cumulative Fluxes per year====
df_cumsum <- df.combined_new %>%
  select(TIMESTAMP, year.x, doy.x, all_of(c(flux_col, model_col, model_v2_col))) %>%
  pivot_longer(
    cols = all_of(c(flux_col, model_col, model_v2_col)),
    names_to = "source",
    values_to = "value"
  ) %>%
  group_by(year.x, source) %>%
  arrange(doy.x, .by_group = TRUE) %>%
  mutate(cumulative_value = cumsum(replace_na(value, 0))) %>%
  ungroup()

for (i in seq(1, length(all_years), by = 4)) {
  years_to_plot <- all_years[i:min(i+3, length(all_years))]
  df_subset <- df_cumsum %>% filter(year.x %in% years_to_plot)

  p_cumsum <- ggplot(df_subset, aes(x = doy.x, y = cumulative_value, color = source)) +
    geom_line(size = 1) +
    facet_wrap(~ year.x, ncol = 2) +
    xlab("DOY") +
    ylab(expression(gC/m^2)) +
    scale_color_manual(
      values = setNames(c("red", "blue", "darkgreen"), c(flux_col, model_col, model_v2_col)),
      labels = c("Flux Tower", "Model", "Model v2")
    ) +
    labs(title = paste0("Cumulative ", compare_name, " Comparison")) +
    my_theme

  print(p_cumsum)
}

# ==== Cumulative Fluxes for all years continuously ====
df_cumsum_all <- df.combined_new %>%
  select(TIMESTAMP, year.x, doy.x, all_of(c(flux_col, model_col, model_v2_col))) %>%
  pivot_longer(
    cols = all_of(c(flux_col, model_col, model_v2_col)),
    names_to = "source",
    values_to = "value"
  ) %>%
  arrange(TIMESTAMP) %>%      # order by full timestamp (chronological)
  group_by(source) %>%
  mutate(cumulative_value = cumsum(replace_na(value, 0))) %>%
  ungroup()

df_cumsum_all <- df_cumsum_all %>%
    mutate(TIMESTAMP = as.POSIXct(TIMESTAMP, format = "%Y-%m-%d"))

# Plot cumulative fluxes continuously for all years
p_cumsum_all <- ggplot(df_cumsum_all, aes(x = TIMESTAMP, y = cumulative_value, color = source, group=source)) +
  geom_line(size = 1) +
  xlab("Time") +
  ylab(expression(gC/m^2)) +
  scale_color_manual(
    values = setNames(c("red", "blue", "darkgreen"), c(flux_col, model_col, model_v2_col)),
    labels = c("Flux Tower", "Model", "Model v2")
  ) +
  labs(title = paste0("Cumulative ", compare_name, " Across Years")) +
  my_theme

print(p_cumsum_all)


```



# PART 4: Evaluating level of agreement
NOTE: this is a simplified approach that evaluates model-data agreement assuming model and flux tower data are from a single point, per year, and over the entire period that data is available for both. This means that there aren't spatial components to the evaluation. 
```{r}
# ===== USER INPUT SECTION ====================================================
nee_flux <- "NEE_VUT_REF"    
nee_model <- "NEE"    
gpp_flux <- "GPP_NT_VUT_REF"
gpp_model <- "GPP" 
reco_flux <- "RECO_NT_VUT_REF"
reco_model <- "ER" 
nee_model_v2 <- "NEE_v2"
gpp_model_v2 <- "GPP_v2"
reco_model_v2 <- "ER_v2"

selected_years <- list(1996:2015) #change years to plot; it can handle several batches (ex: list(2001:2005, 2010:20015))
# =============================================================================

year_range <- unlist(selected_years)


df.combined_eval <- df.combined %>%
  filter(!is.na(NEE_VUT_REF_QC)) %>%
  mutate(
    NEE_flux = .data[[nee_flux]],
    NEE_model = .data[[nee_model]] * 86400,
    NEE_model_v2 = .data[[nee_model_v2]] * 86400,

    GPP_flux = .data[[gpp_flux]],
    GPP_model = .data[[gpp_model]] * 86400,
    GPP_model_v2 = .data[[gpp_model_v2]] * 86400,

    Reco_flux = .data[[reco_flux]],
    Reco_model = .data[[reco_model]] * 86400,
    Reco_model_v2 = .data[[reco_model_v2]] * 86400
  ) %>%
  filter(year.x %in% year_range)


variables_compare <- c(
  "year.x", "month.x", "date.x", "doy.x", "TIMESTAMP",
  "NEE_flux", "NEE_model", "NEE_model_v2",
  "GPP_flux", "GPP_model", "GPP_model_v2",
  "Reco_flux", "Reco_model", "Reco_model_v2"
)

setDT(df.combined_eval)
df.combined_sub <- df.combined_eval[, ..variables_compare]

# Replace negative tower GPP values with NA
df.combined_sub$GPP_flux[df.combined_sub$GPP_flux < 0] <- NA

df.combined_sub <- df.combined_sub %>%
  pivot_longer(
    cols = c(
      NEE_flux, NEE_model, NEE_model_v2,
      GPP_flux, GPP_model, GPP_model_v2,
      Reco_flux, Reco_model, Reco_model_v2
    ),
    names_to = "variable",
    values_to = "value"
  )

df.combined_sub <- df.combined_sub %>%
  mutate(
    data_type = case_when(
      grepl("_flux$", variable) ~ "Tower",
      grepl("_model_v2$", variable) ~ "Model_v2",
      grepl("_model$", variable) ~ "Model",
      TRUE ~ NA_character_
    ),
    flux_type = case_when(
      grepl("^NEE", variable) ~ "NEE",
      grepl("^GPP", variable) ~ "GPP",
      grepl("^Reco", variable) ~ "Reco",
      TRUE ~ NA_character_
    )
  )

df.combined_sub$site_name <- site_name

## filter data

 #filtering out dates where there isn't tower data available for comparison
# Identify dates where Tower data is missing for each flux_type
missing_dates <- df.combined_sub %>%
  filter(data_type == "Tower") %>%
  filter(is.na(value)) %>%
  select(date.x, flux_type)

# Remove all rows for those dates and flux_type combos
df.combined_sub <- df.combined_sub %>%
  anti_join(missing_dates, by = c("date.x", "flux_type"))

## compute interannual and overall metrics, then plot it

# Variables to evaluate
variable_list <- c("GPP", "NEE", "Reco")

all_metrics <- list()  # Initialize list to store results

for (variable_run in variable_list) {
  for (year_val in year_range) {

    obs <- subset(df.combined_sub, data_type == "Tower" & flux_type == variable_run & year.x == year_val)
    model <- subset(df.combined_sub, data_type == "Model" & flux_type == variable_run & year.x == year_val)
    model_v2 <- subset(df.combined_sub, data_type == "Model_v2" & flux_type == variable_run & year.x == year_val)

    # Skip if missing obs
    if (nrow(obs) == 0 || all(is.na(obs$value))) next

    # -------- Evaluate Model vs Tower --------
    if (nrow(model) > 0 && !all(is.na(model$value))) {
      norm_sd <- sd(model$value, na.rm = TRUE) / sd(obs$value, na.rm = TRUE)
      S_corr <- (1 + cor(model$value, obs$value, use = "complete.obs", method = "pearson")) / 2
      bias <- model$value - obs$value
      rmse <- sqrt(mean((model$value - obs$value)^2, na.rm = TRUE))
      e_bias <- ifelse(abs(obs$value) > 0, abs(model$value - obs$value) / abs(obs$value), NA)
      S_bias <- exp(-1 * e_bias)
      S_bias[!is.finite(S_bias)] <- NA
      domain_S_bias <- mean(S_bias, na.rm = TRUE)
      model_anom <- model$value - mean(model$value, na.rm = TRUE)
      obs_anom <- obs$value - mean(obs$value, na.rm = TRUE)
      crmse <- sqrt(mean((model_anom - obs_anom)^2, na.rm = TRUE))
      crms <- sqrt(mean(obs_anom^2, na.rm = TRUE))
      e_rmse <- crmse / crms
      S_rmse <- exp(-1 * e_rmse)
      S <- (domain_S_bias + 2 * S_rmse + S_corr) / 4

      if (!is.na(S)) {
        metric <- c("tower mean", "tower SD", "model mean", "model SD", "SD norm.", "bias", "RMSE",
                    "correlation score", "bias score", "RMSE score", "overall score")
        model_scores <- c(mean(obs$value, na.rm = TRUE), sd(obs$value, na.rm = TRUE),
                          mean(model$value, na.rm = TRUE), sd(model$value, na.rm = TRUE),
                          norm_sd, mean(bias, na.rm = TRUE), rmse,
                          S_corr, domain_S_bias, S_rmse, S)

        score_df <- data.frame(
          site = site_name,
          year = year_val,
          variable = variable_run,
          metric = metric,
          value = model_scores,
          model_version = "Model"
        )

        all_metrics[[paste0(variable_run, "_", year_val, "_Model")]] <- score_df
      }
    }

    # -------- Evaluate Model_v2 vs Tower --------
    if (nrow(model_v2) > 0 && !all(is.na(model_v2$value))) {
      norm_sd <- sd(model_v2$value, na.rm = TRUE) / sd(obs$value, na.rm = TRUE)
      S_corr <- (1 + cor(model_v2$value, obs$value, use = "complete.obs", method = "pearson")) / 2
      bias <- model_v2$value - obs$value
      rmse <- sqrt(mean((model_v2$value - obs$value)^2, na.rm = TRUE))
      e_bias <- ifelse(abs(obs$value) > 0, abs(model_v2$value - obs$value) / abs(obs$value), NA)
      S_bias <- exp(-1 * e_bias)
      S_bias[!is.finite(S_bias)] <- NA
      domain_S_bias <- mean(S_bias, na.rm = TRUE)
      model_anom <- model_v2$value - mean(model_v2$value, na.rm = TRUE)
      obs_anom <- obs$value - mean(obs$value, na.rm = TRUE)
      crmse <- sqrt(mean((model_anom - obs_anom)^2, na.rm = TRUE))
      crms <- sqrt(mean(obs_anom^2, na.rm = TRUE))
      e_rmse <- crmse / crms
      S_rmse <- exp(-1 * e_rmse)
      S <- (domain_S_bias + 2 * S_rmse + S_corr) / 4

      if (!is.na(S)) {
        metric <- c("tower mean", "tower SD", "model mean", "model SD", "SD norm.", "bias", "RMSE",
                    "correlation score", "bias score", "RMSE score", "overall score")
        model_scores <- c(mean(obs$value, na.rm = TRUE), sd(obs$value, na.rm = TRUE),
                          mean(model_v2$value, na.rm = TRUE), sd(model_v2$value, na.rm = TRUE),
                          norm_sd, mean(bias, na.rm = TRUE), rmse,
                          S_corr, domain_S_bias, S_rmse, S)

        score_df <- data.frame(
          site = site_name,
          year = year_val,
          variable = variable_run,
          metric = metric,
          value = model_scores,
          model_version = "Model_v2"
        )

        all_metrics[[paste0(variable_run, "_", year_val, "_Model_v2")]] <- score_df
      }
    }
  }
}


# Combine results from loop and create interannual metric table
metric_df <- do.call(rbind, all_metrics)
row.names(metric_df) <- NULL

print(metric_df) #check out the results!

# Calculate overall metric from yearly results, now including model_version
metric_summary <- metric_df %>%
  group_by(site, variable, metric, model_version) %>%
  summarise(
    value = mean(value, na.rm = TRUE),
    .groups = "drop"
  )

# plot interannual metrics (including model_version)
metric_df_sub <- subset(metric_df, metric %in% c("bias score", "correlation score", 
                                                 "RMSE score", "overall score"))

metric_df_sub$metric <- factor(metric_df_sub$metric, 
                               levels = c("bias score", "correlation score", 
                                          "RMSE score", "overall score"))

# Plot overall metric (including model_version)
scores <- c("bias score", "correlation score", "RMSE score", "overall score")

metric_summary_sub <- subset(metric_summary, metric %in% scores)

# Order metrics nicely
metric_summary_sub$metric <- factor(
  metric_summary_sub$metric,
  levels = c("bias score", "correlation score", "RMSE score", "overall score")
)

# List of variables to plot separately
variables_to_plot <- c("GPP", "NEE", "Reco")

# Plot interannual metrics separately for each variable
walk(variables_to_plot, function(var_name) {
  metric_df_sub_var <- metric_df_sub %>% filter(variable == var_name)
  
  p <- ggplot(metric_df_sub_var, aes(x = metric, y = factor(year), fill = value)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.2f", value)), size = 3) +
    scale_fill_gradientn(colors = brewer.pal(n = 10, name = "Spectral"), limits = c(0, 1)) +
    facet_wrap(~ model_version, scales = "free") +  # facet by model version only
    my_theme +
    labs(title = paste0("Flux vs Model Evaluation per Year - ", var_name),
         x = NULL, y = "Year") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)
})

# Plot overall metrics across all years
ggplot(metric_summary_sub, aes(x = metric, y = site, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f", value)), size = 5, angle = 90) +
  scale_fill_gradientn(colors = brewer.pal(n = 10, name = "Spectral"), limits = c(0, 1), name = "Score") +
  facet_grid(variable ~ model_version, scales = "free") +  # facet by variable (rows) and model_version (columns)
  my_theme +
  labs(title = "Flux vs Model Evaluation Across All Years",
       x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))





#plot_save <- "model_scoring_GPP_NEE.png"
#ggsave(filename = plot_save, dpi = 500)

```





# Optional: evaluate flux vs model per month (across years); are there months when the model and flux tower agree more?
```{r}
# Variables to evaluate
variable_list <- c("GPP", "NEE", "Reco")
all_monthly_metrics <- list()  # to store results

months_to_evaluate <- sort(unique(df.combined_sub$month.x))

for (variable_run in variable_list) {
  for (month_val in months_to_evaluate) {
    
    # Subset observation data (Tower)
    obs <- subset(df.combined_sub, data_type == "Tower" & flux_type == variable_run & month.x == month_val)

    # Skip if obs missing
    if (nrow(obs) == 0 || all(is.na(obs$value))) next

    # Subset model and model_v2 data
    model    <- subset(df.combined_sub, data_type == "Model"    & flux_type == variable_run & month.x == month_val)
    model_v2 <- subset(df.combined_sub, data_type == "Model_v2" & flux_type == variable_run & month.x == month_val)
    
    # ---- Function to calculate metrics ----
    calc_metrics <- function(model_data, obs_data, site_name, variable_run, month_val, model_version_label) {
      if (nrow(model_data) == 0 || all(is.na(model_data$value))) return(NULL)
      
      norm_sd <- sd(model_data$value, na.rm = TRUE) / sd(obs_data$value, na.rm = TRUE)
      S_corr  <- (1 + cor(model_data$value, obs_data$value, use = "complete.obs", method = "pearson")) / 2
      bias    <- model_data$value - obs_data$value
      rmse    <- sqrt(mean((model_data$value - obs_data$value)^2, na.rm = TRUE))
      
      e_bias  <- ifelse(abs(obs_data$value) > 0, abs(model_data$value - obs_data$value) / abs(obs_data$value), NA)
      S_bias  <- exp(-1 * e_bias)
      S_bias[!is.finite(S_bias)] <- NA
      domain_S_bias <- mean(S_bias, na.rm = TRUE)
      
      model_anom <- model_data$value - mean(model_data$value, na.rm = TRUE)
      obs_anom   <- obs_data$value   - mean(obs_data$value, na.rm = TRUE)
      crmse      <- sqrt(mean((model_anom - obs_anom)^2, na.rm = TRUE))
      crms       <- sqrt(mean(obs_anom^2, na.rm = TRUE))
      e_rmse     <- crmse / crms
      S_rmse     <- exp(-1 * e_rmse)
      
      S <- (domain_S_bias + 2 * S_rmse + S_corr) / 4
      
      metric <- c("tower mean", "tower SD", "model mean", "model SD", "SD norm.", "bias", "RMSE",
                  "correlation score", "bias score", "RMSE score", "overall score")
      
      model_scores <- c(
        mean(obs_data$value, na.rm = TRUE), sd(obs_data$value, na.rm = TRUE),
        mean(model_data$value, na.rm = TRUE), sd(model_data$value, na.rm = TRUE),
        norm_sd, mean(bias, na.rm = TRUE), rmse,
        S_corr, domain_S_bias, S_rmse, S
      )
      
      data.frame(site = site_name, month = month_val, variable = variable_run,
                 metric = metric, value = model_scores, model_version = model_version_label)
    }
    
    # Calculate metrics for Model
    res_model <- calc_metrics(model, obs, site_name, variable_run, month_val, "Model")
    # Calculate metrics for Model_v2
    res_model_v2 <- calc_metrics(model_v2, obs, site_name, variable_run, month_val, "Model_v2")
    
    # Append results to list if not NULL
    if (!is.null(res_model)) all_monthly_metrics[[paste0(variable_run, "_", month_val, "_Model")]] <- res_model
    if (!is.null(res_model_v2)) all_monthly_metrics[[paste0(variable_run, "_", month_val, "_Model_v2")]] <- res_model_v2
  }
}

# Combine into final table
monthly_metric_df <- do.call(rbind, all_monthly_metrics)
row.names(monthly_metric_df) <- NULL

# Subset to plot only key scoring metrics
scores <- c("bias score", "correlation score", "RMSE score", "overall score")

monthly_metric_plot_df <- subset(monthly_metric_df, metric %in% scores)

# Order metrics
monthly_metric_plot_df$metric <- factor(
  monthly_metric_plot_df$metric,
  levels = c("bias score", "correlation score", "RMSE score", "overall score")
)

# Add month name labels if needed
monthly_metric_plot_df$month <- factor(monthly_metric_plot_df$month, 
                                       levels = 1:12, 
                                       labels = month.abb)

# Plot monthly metrics across all years
variables_to_plot <- c("GPP", "NEE", "Reco")

walk(variables_to_plot, function(var_name) {
  df_plot <- monthly_metric_plot_df %>% filter(variable == var_name)
  
  p <- ggplot(df_plot, aes(x = metric, y = month, fill = value)) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.2f", value)), size = 4) +
    scale_fill_gradientn(colors = brewer.pal(10, "Spectral"), limits = c(0, 1), name = "Score") +
    facet_wrap(~ model_version, scales = "free") +
    scale_y_discrete(limits = rev(levels(df_plot$month))) +  # Reverse month order: Jan on top
    my_theme +
    labs(title = paste0("Model Performance by Month Across Years - ", var_name),
         x = NULL, y = NULL) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  print(p)
})




```


